\section{Reproduction and Additional Details}
\label{sec:appendix:exp}

\subsection{Workload Reproduction}
\label{sec:appendix:repro}

All seven workloads are reproduced using the commands below.
Each workload requires an NVIDIA GPU (Ampere architecture minimum; all experiments in this paper
use an A100-SXM4-80GB), \nsys $\ge$ 2025.1, \ncu $\ge$ 2025.4.1, PyTorch 2.5+, and CUDA 12.x.

\begin{lstlisting}[style=pystyle]
# Install
pip install -e .

# Phase 1: correlation pass (no nsys, builds torch.profiler attribution map)
python nvidia/scripts/run_workload.py \
    --workload examples/gpt2/workload.py \
    --output-prefix runs/gpt2/baseline \
    --inductor-debug-dir runs/gpt2/inductor_debug \
    --correlation-pass

# Phase 2: NVTX capture under nsys
nsys profile --trace=cuda,nvtx --output=runs/gpt2/baseline \
    python nvidia/scripts/run_workload.py \
        --workload examples/gpt2/workload.py \
        --output-prefix runs/gpt2/baseline \
        --inductor-debug-dir runs/gpt2/inductor_debug

# Build manifest and collect hardware counters (requires sudo for ncu)
python -m nvidia.operator_profiler.cli manifest \
    --prefix runs/gpt2/baseline \
    --output runs/gpt2/baseline.manifest.json

python -m nvidia.operator_profiler.cli map \
    --manifest runs/gpt2/baseline.manifest.json \
    --workload examples/gpt2/workload.py \
    --ncu-sudo --output runs/gpt2/profile.json
\end{lstlisting}

Profiling results, optimized workload scripts, and \texttt{profile.json} /
\texttt{profile\_optimized.json} pairs for measured examples are in
\texttt{examples/} in the repository.  Each subdirectory contains
\texttt{OPTIMIZED\_WORKLOAD.md} with the measured before-and-after hardware counter values.
Layer deduplication is activated automatically by \texttt{run\_workload.py} when structurally
repeated partitions are detected; the partition equivalence map is written to
\texttt{<prefix>.part.json} and passed to the ncu orchestrator to skip duplicate-partition
kernel replays.

\subsection{Hardware Counter Reference}
\label{sec:appendix:counters}

\tabref{tab:counters_full} lists the architecture-specific \ncu counter name
variants for the five counters that differ across Ampere, Hopper, and
Blackwell.  \sys queries all variants and uses the first that returns a
non-null value.

\begin{table}[h]
\centering
\caption{Architecture-specific \ncu counter name fallbacks used by \sys.}
\label{tab:counters_full}
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{p{3.2cm}p{4.5cm}p{5cm}}
\toprule
\textbf{Profile Field} & \textbf{Ampere name} & \textbf{Hopper / Blackwell name} \\
\midrule
\texttt{dram\_bytes\_read}
  & \texttt{dram\_\_bytes\_read.sum}
  & \texttt{fbpa\_\_dram\_read\_bytes.sum} \\
\texttt{dram\_bytes\_written}
  & \texttt{dram\_\_bytes\_write.sum}
  & \texttt{fbpa\_\_dram\_write\_bytes.sum} \\
\texttt{dram\_throughput\_pct}
  & \texttt{dram\_\_throughput.avg.pct\_of\_peak\_sustained\_elapsed}
  & \texttt{gpu\_\_dram\_throughput.avg.pct\_of\_peak\_sustained\_elapsed} \\
\texttt{sm\_throughput\_pct}
  & \texttt{sm\_\_throughput.avg.pct\_of\_peak\_sustained\_elapsed}
  & same \\
\texttt{achieved\_occupancy}
  & \texttt{sm\_\_warps\_active.avg.pct\_of\_peak\_sustained\_active}
  & same \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Additional Limitations}
\label{sec:appendix:limitations}

Beyond the limitations noted in \secref{sec:conclusion}:
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{\ncu replay timing inflation.}  All \texttt{duration\_ns} values in
    \texttt{profile.json} reflect \ncu's serialized replay execution, which runs
    2--5$\times$ slower than real wall-clock execution.  Values should be used for
    relative comparison only; \texttt{torch.cuda.Event} timing is authoritative for
    absolute latency.
  \item \textbf{Multi-GPU attribution.}  The current pipeline profiles a single GPU.
    Distributed workloads with NCCL communication require per-stream, per-device
    profiling runs; cross-device attribution is not yet supported.
  \item \textbf{Dynamic shapes.}  Shape validation enforces that \ncu replay inputs
    match the \nsys capture.  Models with truly dynamic shapes (varying sequence
    length per call) require separate captures per shape bucket.
\end{itemize}