\section{Background}
\label{sec:background}

\subsection{GPU Execution Model and Clock Domains}

A PyTorch forward pass translates a sequence of \texttt{aten::} operator calls through the
dispatcher into one or more CUDA kernel launches.  Each kernel executes on a set of Streaming
Multiprocessors (SMs), organized into thread blocks (CTAs); each CTA contains multiple warps of
32 threads.  GPU hardware exposes performance counters via CUPTI (CUDA Profiling Tools Interface),
which operates in a hardware-domain clock derived from the GPU's free-running oscillator.

\textbf{The clock domain problem.}  Two GPU profiling tools observe the same kernel execution
through different timing mechanisms.  Nsight Systems (\nsys) reads CUPTI activity records that
carry GPU-domain timestamps (\texttt{start\_ns}, \texttt{end\_ns}) from
\texttt{CUPTI\_ACTIVITY\_KIND\_KERNEL}.  Nsight Compute (\ncu) replays each kernel in a separate
profiling pass, injecting performance counter reads between kernel invocations; its internal
timestamps reflect the replay-mode execution clock, not the original capture clock.  Even on the
same physical GPU, the two clocks diverge by tens of microseconds per kernel due to replay overhead
and kernel serialization.  \textbf{Timestamp joins across these two tools are incorrect and
introduce systematic misattribution.}  \sys avoids this entirely (\secref{sec:attribution}).

NVTX (NVIDIA Tools Extension) ranges are CPU-side annotations pushed and popped by the host
process.  PyTorch's \texttt{torch.autograd.profiler.emit\_nvtx()} context manager inserts
\texttt{nvtxRangePushA("aten::linear")} before and \texttt{nvtxRangePop()} after every
\texttt{aten::} dispatch call, labeling the host-side window during which that operator's CUDA
kernels are launched.  These annotations appear in the \nsys SQLite export as
\texttt{NVTX\_EVENTS} rows with CPU-domain timestamps.

\subsection{PyTorch Compilation Stack}

PyTorch 2.x introduced TorchDynamo and TorchInductor \cite{pytorch2024} as a two-stage compilation
pipeline.  Dynamo intercepts Python bytecode, identifies subgraphs of PyTorch operations, and
traces them to an FX graph in the Aten IR (\texttt{torch.ops.aten.*}).  Inductor then lowers this
graph to Triton-generated GPU kernels, applying fusion heuristics such as pointwise-into-reduction
and elementwise-chaining.

Custom compilation backends are registered via:
\begin{lstlisting}[style=pystyle]
@torch._dynamo.register_backend("my_backend")
def my_backend(gm: torch.fx.GraphModule,
               example_inputs) -> Callable:
    # apply FX passes, then compile
    return torch._inductor.compile_fx(gm, example_inputs)
\end{lstlisting}
This hook intercepts the Aten IR \texttt{GraphModule} before Inductor processes it, enabling
hardware-informed FX passes to modify the graph based on profiling data (\secref{sec:pgo}).

Triton \cite{triton2019} generates fused kernels whose names encode constituent operations, \eg
\texttt{triton\_poi\_fused\_relu\_addmm\_0} for a pointwise kernel fusing \texttt{relu} and
\texttt{addmm}.  When NVTX ranges are absent or ambiguous, these names provide a low-confidence
attribution signal (\secref{sec:heuristic}).

\subsection{Roofline Bottleneck Taxonomy}
\label{sec:roofline}

The Roofline model \cite{williams2009roofline} characterizes a kernel's performance ceiling as
$\min(\text{peak FLOP/s},\ \text{AI} \times \text{peak bandwidth})$, where arithmetic intensity
$\text{AI} = \text{FLOPs} / \text{bytes}$.  A kernel below the bandwidth ceiling is
\emph{memory-bound}; above the ridge point it is \emph{compute-bound}.  \sys operationalizes this
taxonomy at the per-operator level using hardware counters:

\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Tensor Core idle} (\texttt{tensor\_core\_active\_pct} $\approx 0$):
    compute-bound path is not engaging hardware-accelerated units; fix via dtype change
    (\secref{sec:passes}).
  \item \textbf{Wave starvation} (\texttt{achieved\_occupancy} $< 20\%$, Waves/SM $< 1$):
    insufficient parallelism to fill the GPU; fix via batch padding or kernel fusion
    (\secref{sec:passes}).
  \item \textbf{Register pressure} (\texttt{registers\_per\_thread} $> 128$):
    limits concurrent warps; often resolves with dtype change reducing per-thread state.
  \item \textbf{Cache miss pressure} (\texttt{l2\_hit\_rate} $< 30\%$ on GEMM paths):
    layout or access-pattern mismatch; fix via weight pre-transposition
    (\secref{sec:passes}).
  \item \textbf{Memory spill} (\texttt{local\_memory\_spills} $> 0$):
    register-file overflow to L1; fix via kernel substitution (FlashAttention-2 vs.\ standard FMHA,
    \secref{sec:passes}).
\end{itemize}

These five patterns cover the bottlenecks observed across all six workloads evaluated in
\secref{sec:experiments}.  \figref{fig:roofline} maps them onto the roofline model.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{loglogaxis}[
  width=0.82\columnwidth,
  height=5.8cm,
  xlabel={Arithmetic Intensity (FLOP/byte)},
  ylabel={Performance (GFLOP/s)},
  xlabel style={font=\small},
  ylabel style={font=\small},
  tick label style={font=\scriptsize},
  xmin=0.05, xmax=150,
  ymin=1,    ymax=4000,
  grid=both,
  grid style={gray!20},
  legend style={
    at={(0.97,0.03)}, anchor=south east,
    font=\scriptsize, draw=gray!50,
    fill=white, fill opacity=0.85,
    row sep=-1pt,
  },
  clip=false,
]

%% ---- Memory bandwidth ceiling (slope = HBM BW * AI, Blackwell HBM3e ~1457 GB/s) ----
%% Represented schematically: slope line y = BW * x, ridge at ~x=2 (arbitrary schematic units)
\addplot[thick, blue!60!black, domain=0.05:2.5] {500 * x};

%% ---- FP32 scalar compute ceiling ----
\addplot[thick, orange!80!black, dashed, domain=2.5:150] coordinates {(2.5,1250) (150,1250)};
\node[font=\scriptsize, text=orange!80!black, anchor=west] at (axis cs:3,1450)
  {FP32 SIMT ceiling};

%% ---- BF16 Tensor Core ceiling (8x higher) ----
\addplot[thick, green!60!black, domain=0.3:150] coordinates {(0.3,2400) (150,2400)};
\node[font=\scriptsize, text=green!60!black, anchor=west] at (axis cs:3,2700)
  {BF16 Tensor Core ceiling ($8\times$)};

%% ---- Ridge point markers ----
\addplot[gray!50, dotted, thin] coordinates {(2.5,1) (2.5,1250)};  % FP32 ridge
\addplot[gray!50, dotted, thin] coordinates {(0.3,1) (0.3,2400)};  % BF16 ridge (schematic)

%% ---- Baseline operator positions (schematic dots) ----
% Wave starvation: low AI, very low perf
\addplot[only marks, mark=*, mark size=3pt, red!70!black]
  coordinates {(0.08,4) (0.12,6) (0.15,5)};
\node[font=\scriptsize, text=red!70!black, align=center, anchor=north]
  at (axis cs:0.1,3.2) {\textbf{Wave}\\\textbf{starvation}\\Waves/SM$<$1};

% TC idle: moderate AI, stuck below FP32 ceiling
\addplot[only marks, mark=square*, mark size=2.8pt, orange!80!black]
  coordinates {(5,180) (10,250) (20,300)};
\node[font=\scriptsize, text=orange!80!black, align=center, anchor=south]
  at (axis cs:10,170) {\textbf{TC idle}\\TC\_pct$\approx$0};

% Register/spill pressure: moderate AI, moderate perf (below compute ceiling)
\addplot[only marks, mark=triangle*, mark size=3pt, purple!70!black]
  coordinates {(3,650) (4,700)};
\node[font=\scriptsize, text=purple!70!black, align=center, anchor=south west]
  at (axis cs:4.2,680) {\textbf{Reg./spill}\\pressure};

% Cache miss: BW-bound but underperforming slope
\addplot[only marks, mark=diamond*, mark size=3pt, brown!70!black]
  coordinates {(0.9,200) (1.3,270)};
\node[font=\scriptsize, text=brown!70!black, align=center, anchor=north]
  at (axis cs:1.1,170) {\textbf{Cache miss}\\L2 hit$<$30\%};

%% ---- Arrows showing optimized positions ----
\draw[-{Stealth[length=5pt]}, thick, green!50!black, dashed]
  (axis cs:0.1,5.5) -- (axis cs:0.3,60)
  node[midway, right=2pt, font=\scriptsize, text=green!50!black] {QKV fusion};

\draw[-{Stealth[length=5pt]}, thick, green!50!black, dashed]
  (axis cs:8,240) -- (axis cs:8,2100)
  node[midway, right=2pt, font=\scriptsize, text=green!50!black] {BF16};

\draw[-{Stealth[length=5pt]}, thick, green!50!black, dashed]
  (axis cs:3.5,680) -- (axis cs:3.5,1900)
  node[midway, right=2pt, font=\scriptsize, text=green!50!black] {FlashAttn};

\addlegendentry{Memory BW ceiling}
\addlegendentry{FP32 SIMT ceiling}
\addlegendentry{BF16 TC ceiling}

\end{loglogaxis}
\end{tikzpicture}
\caption{\textbf{Roofline bottleneck taxonomy operationalized by \sys} (schematic, log-log scale).
  Each labeled region corresponds to one hardware counter pattern: \textcolor{red!70!black}{wave starvation}
  (Waves/SM~$<$~1), \textcolor{orange!80!black}{Tensor Core idle} (\texttt{tensor\_core\_active\_pct}~$\approx$~0),
  \textcolor{purple!70!black}{register/spill pressure} (\texttt{registers\_per\_thread}~$>$~128 or
  \texttt{local\_memory\_spills}~$>$~0), and \textcolor{brown!70!black}{cache miss pressure}
  (\texttt{l2\_hit\_rate}~$<$~30\%).  Dashed arrows show the direction each \sys FX pass moves
  the operator.  Ceiling values and operator positions are schematic; see
  \tabref{tab:hwcounters} for measured values.}
\label{fig:roofline}
\end{figure}
