\section{Conclusion}
\label{sec:conclusion}

We presented \sys, a hardware-attributed GPU profiling and profile-guided optimization system that
bridges the gap between PyTorch's operator-level abstractions and NVIDIA's kernel-level hardware
performance counters. Our contributions---an attribution pipeline with per-stream interval trees,
invocation-order matching, and Inductor debug-artifact enrichment for compiled models; a 20-counter
curated metric set with duration-weighted aggregation and architecture-specific fallbacks;
\texttt{FxPassRunner} and \texttt{UniqueSubgraphRegistry} for pattern-based FX graph optimization
of unique layer representatives with hardware-evidence-motivated patterns; layer deduplication that
reduces \ncu replay time by $O(\text{num\_duplicate\_layers})$; a closed-loop validation protocol;
and a Claude Code plugin that automates the full eight-stage optimization workflow via a single
\texttt{/optimize~workload.py} command---collectively enable an automated, evidence-driven
optimization loop at the intersection of PyTorch operator profiling and hardware counter measurement.

The key results are not the speedup numbers themselves, but what they are built from: every
optimization decision is grounded in a specific hardware counter reading (e.g.,
\texttt{tensor\_core\_active\_pct = 0\%} $\to$ apply BF16 promotion achieving 2.93$\times$ on
ConvBlock and ${\sim}$2.5--3.0$\times$ on GPT-2 on NVIDIA A100-SXM4-80GB;
\texttt{local\_memory\_spills = 24,468,480\,B} $\to$ BF16 fmha eliminates spills entirely), and
every claimed improvement is verified by re-profiling with the same hardware counter pipeline.
This makes optimizations reproducible, auditable, and non-trivially distinct from rule-of-thumb
best practices: some bottlenecks (register spill in FMHA, NHWC layout coercion on every forward
pass) are invisible to wall-clock profilers and require hardware counter attribution to diagnose.

\paragraph{Limitations.}
Attribution coverage degrades for cuDNN-backed operators (\texttt{nn.LSTM},
\texttt{nn.MultiheadAttention} under cuDNN) where internal kernel names carry no NVTX
correspondence and no Inductor debug metadata is available; expected unattributed rates of
80--90\% make hardware-counter-driven optimization inapplicable, and bottleneck diagnosis must
fall back to kernel-name inference.  \ncu serializes kernels for replay, making concurrent
multi-stream profiles require separate per-stream runs.  Optimization patterns are currently
user-defined per-workload (implemented in each \texttt{*\_optimized.py}); automatic pattern
selection from counter evidence---via learned thresholds or a rule engine over the
bottleneck-to-pattern mapping---is left as future work.

\paragraph{Future work.}
Online profiling via CUPTI Perfworks streaming would eliminate the separate \ncu pass, enabling
profiling during ordinary training without a dedicated replay run.  Multi-GPU attribution with
NCCL stream synchronization events would extend the pipeline to distributed workloads.
Counter-guided neural architecture search---using per-operator hardware utilization as a fitness
signal rather than validation accuracy alone---represents a longer-term direction for hardware-aware
model design.  Finally, extending the counter name mapping and pass library to AMD ROCm (\texttt{rocprof}
+ \texttt{rocm-smi}) would generalize \sys beyond NVIDIA hardware.
