\section{Introduction}
\label{sec:intro}

The throughput of a GPU workload is determined by a handful of hardware bottlenecks: memory
bandwidth saturation, Tensor Core under-utilization, register-pressure-limited occupancy, or wave
starvation from insufficient parallelism. NVIDIA provides tools to measure each---Nsight Compute
(\ncu) collects over 90 hardware counters per kernel invocation---yet the dominant optimization
workflow for PyTorch practitioners remains largely manual: profile timing in \texttt{torch.profiler},
identify the slow operator, guess at the cause, edit the model, repeat. The bottleneck taxonomy
(compute-bound vs.\ memory-bound vs.\ occupancy-limited) that Williams \etal
\cite{williams2009roofline} formalized for HPC remains difficult to apply at the operator level in
PyTorch, not because the hardware data is absent, but because nothing bridges it to the framework.

\paragraph{The attribution gap.}
\texttt{torch.profiler} and Kineto \cite{kineto2021} record wall-clock durations for each
\texttt{aten::} operator, but do not access hardware counters. \ncu collects per-kernel hardware
metrics---occupancy, DRAM bytes, Tensor Core utilization, cache hit rates---but its output is keyed
on internal CUDA kernel names such as
\texttt{ampere\_sgemm\_128x32\_tn} or \texttt{triton\_poi\_fused\_relu\_addmm\_0}, with no awareness
of which PyTorch operator dispatched them. Bridging the two requires matching records across two
profiling tools that run in separate passes, use incompatible clock domains (CUPTI GPU timestamps
vs.\ \ncu injection-mode timestamps), and produce outputs with no shared stable key.

\paragraph{Why existing solutions fall short.}
The manual bridging workflow---run \nsys to record the kernel timeline, run \ncu for hardware
counters, manually correlate by kernel name---is $O(\text{operators})$ human effort per model
variant, breaks completely for Triton-fused kernels (whose names encode fused operations as opaque
hash suffixes), and does not compose with the feedback loop needed for profile-guided optimization.
TorchInductor \cite{pytorch2024} applies graph-level fusion heuristically, without hardware counter
feedback; the question ``did this fusion actually improve Tensor Core utilization?'' has no
programmatic answer in the standard toolchain.

\paragraph{Our approach.}
We observe that reliable attribution does not require timestamp matching.  Because \ncu replays
kernels in the same execution order they appear in the \nsys trace, a kernel's position within a
\texttt{(kernel\_name, invocation\_index)} sequence is a stable identifier that is clock-immune.
For operator-level attribution, the NVTX ranges that \texttt{torch.autograd.profiler.emit\_nvtx}
pushes around every \texttt{aten::} dispatch form a temporal bracket on the host CPU timeline;
per-stream interval trees over these ranges attribute each kernel to its enclosing operator in
$O(\log n)$ per query. For Triton-fused kernels with no NVTX bracket, a regex parser extracts the
constituent operator names from the mangled kernel identifier.  \figref{fig:overview} shows the
complete end-to-end pipeline.

\paragraph{Contributions.}
We present \textbf{\sys}, a hardware-attributed GPU profiling and profile-guided optimization
system for PyTorch. Our contributions are:

\begin{enumerate}

\item \textbf{Attribution pipeline (\secref{sec:attribution}).}  A
  \texttt{StreamIntervalTree} keyed on \texttt{(stream\_id, device\_id)} pairs attributes each
  CUDA kernel to an \texttt{aten::} operator via NVTX temporal enclosure (MEDIUM confidence). A
  torch.profiler invocation-order join provides HIGH-confidence attribution for registered operators.
  An Inductor debug-artifact enrichment pass recovers ground-truth fused operator identity from
  \texttt{output\_code.py} comments, upgrading previously unattributed Inductor-compiled kernels to
  MEDIUM confidence. Kernels matching neither method are stored in \texttt{unattributed\_kernels[]}
  and never silently dropped. Invocation-order matching between \nsys and \ncu avoids clock-domain
  joins. Eight edge cases---multi-stream execution, CUDA Graph replay, JIT warm-up kernels, async
  launches, dynamic shapes, and fused kernels---are handled explicitly.

\item \textbf{20-counter curated metric set (\secref{sec:metrics}).}  Twenty hardware counters
  spanning all bottleneck axes are collected with architecture-specific name fallbacks for Ampere,
  Hopper, and Blackwell. Duration-weighted aggregation ensures that long-running kernels dominate
  operator-level summaries when multiple kernels are attributed to one operator.

\item \textbf{Custom Torch Dynamo PGO backend (\secref{sec:pgo}).}  A \texttt{@register\_backend}
  hook inserts six hardware-evidence-motivated FX graph passes at the Aten IR level before Inductor
  lowering: BF16 casting, QKV projection fusion, FlashAttention substitution, weight
  pre-transposition, activation normalization, and batch padding.  Each pass is triggered only when
  the profiler diagnoses the specific counter symptom that motivates it.

\item \textbf{Closed-loop validation (\secref{sec:experiments}).}  A four-step protocol---profile
  $\to$ diagnose $\to$ transform $\to$ re-profile---verifies that hardware bottlenecks are
  eliminated, not merely shifted.  Per-operator hardware counter deltas confirm systematic
  improvement across all six evaluated workloads.

\end{enumerate}

\paragraph{Results.}
On a TransformerBlock (hidden dim 512, 8 heads, FFN ratio $4\times$) running on an NVIDIA RTX PRO
6000 Blackwell, \sys achieves \textbf{6.3$\times$} per-sample speedup over the eager-mode FP32
baseline. Tensor Core utilization rises from 0\% to ${\sim}85\%$; register spills from the FMHA
kernel drop from 757,760 to zero; Waves/SM on the Q/K/V projections rises from 0.11 to 0.68.
Across six workloads on two GPU architectures, end-to-end speedups range from 1.8$\times$
(cuDNN LSTM, opaque kernel path) to 6.3$\times$ (TransformerBlock), each verified by hardware
counter before-and-after comparison.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\small,
  box/.style={rounded corners=3pt, draw, align=center, minimum height=1.8em, inner sep=5pt},
  toolbox/.style={box, fill=blue!12, draw=blue!50!black},
  attrbox/.style={box, fill=orange!15, draw=orange!70!black},
  outbox/.style={box, fill=green!12, draw=green!50!black},
  pgobox/.style={box, fill=red!10, draw=red!60!black},
  arr/.style={-{Stealth[length=5pt]}, thick},
  darr/.style={-{Stealth[length=5pt]}, thick, dashed, gray},
  lbl/.style={font=\scriptsize\itshape, text=gray!70!black},
]

%% ---- Row 1: Profiling stage ----
\node[toolbox] (workload) at (0,0)   {\texttt{workload.py}};
\node[toolbox] (nsys)     at (2.7,0) {\nsys capture\\[-1pt]\texttt{--trace=cuda,nvtx}};
\node[toolbox] (sqlite)   at (5.5,0) {SQLite\\[-1pt]export};
\node[toolbox] (ncu)      at (8.2,0) {\ncu replay\\[-1pt](per kernel)};

\draw[arr] (workload) -- (nsys);
\draw[arr] (nsys)     -- (sqlite);
\draw[arr] (sqlite)   -- (ncu);

\node[lbl, above=2pt of nsys]   {Stage A};
\node[lbl, above=2pt of ncu]    {Stage B};

%% ---- Row 2: Attribution engine ----
\node[attrbox] (iom)    at (2.7,-1.65) {Invocation-Order\\[-1pt]Matching};
\node[attrbox] (sit)    at (5.5,-1.65) {StreamIntervalTree\\[-1pt]NVTX enclosure};
\node[outbox]  (prof)   at (8.2,-1.65) {\texttt{profile.json}\\[-1pt](20 counters)};

\draw[arr] (sqlite) -- (iom);
\draw[arr] (ncu)    -- (iom);
\draw[arr] (iom)    -- (sit);
\draw[arr] (sit)    -- (prof);

\node[lbl, left=3pt of iom] {\S\ref{sec:attribution}};

%% ---- Row 3: PGO ----
\node[pgobox] (diag)    at (2.7,-3.3)  {Diagnosis\\[-1pt]Engine};
\node[pgobox] (passes)  at (5.5,-3.3)  {6 FX Passes\\[-1pt](Aten IR)};
\node[pgobox] (compile) at (8.2,-3.3)  {\texttt{torch.compile}\\[-1pt]+ Inductor};
\node[outbox] (opt)     at (10.8,-3.3) {Optimized\\[-1pt]model};

\draw[arr] (prof)    -- (diag);
\draw[arr] (diag)    -- (passes);
\draw[arr] (passes)  -- (compile);
\draw[arr] (compile) -- (opt);

\node[lbl, left=3pt of diag]   {\S\ref{sec:pgo}};
\node[lbl, above=2pt of passes] {BF16 · QKV fusion · FlashAttn · pretranspose · activ.\ norm.\ · batch pad};

%% ---- Feedback loop ----
\draw[darr] (opt.south) -- ++(0,-0.45) -| (workload.south)
  node[midway, below, lbl] {closed-loop validation (\S\ref{sec:validation})};

%% ---- Stage labels on left ----
\node[lbl, rotate=90, anchor=center] at (-1.5, 0)     {Profiling};
\node[lbl, rotate=90, anchor=center] at (-1.5, -1.65) {Attribution};
\node[lbl, rotate=90, anchor=center] at (-1.5, -3.3)  {Optimization};

\end{tikzpicture}
\caption{\textbf{\sys system overview.}  Two profiling stages produce \texttt{profile.json}: \nsys
  captures the kernel timeline (Stage~A) and \ncu replays each kernel for hardware counters
  (Stage~B).  The attribution engine bridges the two tools via invocation-order matching and
  per-stream NVTX interval trees.  The PGO backend applies six hardware-evidence-motivated FX graph
  passes before Inductor compilation.  A dashed feedback loop denotes closed-loop validation via
  re-profiling.}
\label{fig:overview}
\end{figure}
