\section{Attribution Pipeline}
\label{sec:attribution}

\sys ingests two data sources produced for the same workload run: (1) an \nsys SQLite export
containing CUDA kernel activity and NVTX range annotations, and (2) \ncu CSV output containing
per-kernel hardware counter measurements. The pipeline produces a hardware-attributed
\texttt{OperatorProfile}: a JSON document mapping each \texttt{aten::} operator to its constituent
kernels, their hardware metrics, and the confidence tier of each attribution.

\subsection{Data Ingestion}
\label{sec:ingestion}

\textbf{Nsight Systems export.}  Running the workload under
\texttt{nsys profile --trace=cuda,nvtx} produces a \texttt{.nsys-rep} file, exported to SQLite
via \texttt{nsys export --type=sqlite}. \sys queries three tables:

\begin{lstlisting}[style=sqlstyle,caption={Kernel and NVTX range extraction from the \nsys SQLite export.}]
-- CUDA kernel activity (GPU-domain timestamps)
SELECT k.correlationId, s.value AS kernel_name,
       k.start AS start_ns, k.end AS end_ns,
       k.streamId, k.deviceId,
       k.gridX, k.gridY, k.gridZ,
       k.blockX, k.blockY, k.blockZ,
       r.globalTid AS host_tid,
       r.start      AS cpu_launch_start_ns
FROM   CUPTI_ACTIVITY_KIND_KERNEL k
JOIN   StringIds s ON s.id = k.shortName
LEFT JOIN CUPTI_ACTIVITY_KIND_RUNTIME r
       ON r.correlationId = k.correlationId
ORDER  BY k.start ASC;

-- NVTX range annotations (CPU-domain timestamps)
SELECT n.text, n.start AS start_ns, n.end AS end_ns,
       n.nestingLevel, n.globalTid AS stream_id
FROM   NVTX_EVENTS n
WHERE  n.end IS NOT NULL
ORDER  BY n.start ASC;
\end{lstlisting}

The \texttt{cpu\_launch\_start\_ns} field from the RUNTIME table is the host-side timestamp when
the CUDA kernel launch API returned, providing a CPU-domain anchor for NVTX bracket queries
without conflating it with the GPU-domain kernel execution timestamp (\secref{sec:clockdomain}).

\textbf{Nsight Compute export.}  For the full set of kernels in the \nsys trace, \sys invokes
a single application-mode replay pass per counter group (4--8 passes total), collecting all
20 metrics simultaneously:
\begin{lstlisting}[style=pystyle]
ncu --replay-mode application \
    --metrics <comma_separated_list> \
    --export <output.ncu-rep> \
    [--kernel-name <filter>] \
    <replay_script.py>
\end{lstlisting}
Application-mode replays the workload once per counter group regardless of the number of
unique kernel names, making profiling cost $O(\text{counter groups})$ rather than
$O(\text{unique kernels})$---a 10--50$\times$ efficiency gain for multi-layer models.
The optional \texttt{--kernel-name} flag filters to a specific kernel for targeted
re-profiling.  The resulting \texttt{.ncu-rep} is exported to CSV
(\texttt{ncu --import --csv}), producing one row per kernel invocation per metric.

\subsection{The Clock Domain Problem and Invocation-Order Matching}
\label{sec:clockdomain}

The fundamental obstacle in cross-tool profiling is that \nsys and \ncu observe the same kernel
execution through incompatible timing mechanisms. \nsys records CUPTI GPU-domain timestamps from
the hardware's free-running clock. \ncu replays kernels in an isolated profiling pass with injected
performance counter reads; its internal invocation clock reflects replay-mode execution, not the
original capture clock. On a single GPU, these two clocks routinely diverge by 10--100~$\mu$s per
kernel due to replay serialization and PMU injection overhead. A naive timestamp join would
incorrectly associate hardware counter measurements with the wrong operator---especially dangerous
for operators with similar execution times.

\textbf{Invocation-Order Matching (Algorithm~\ref{alg:iom}).}  The key insight is that \ncu
replays kernels in the same execution order they appear in the \nsys timeline: the $i$-th
invocation of kernel $K$ in the \nsys trace corresponds to the $i$-th row for kernel $K$ in the
\ncu CSV output.  We exploit this ordering to assign hardware metrics without any timestamp
comparison:

\begin{algorithm}[t]
\caption{Invocation-Order Matching (\nsys $\times$ \ncu)}
\label{alg:iom}
\begin{algorithmic}[1]
\Require \texttt{nsys\_kernels}: list of \texttt{KernelRow} sorted by \texttt{start\_ns} per stream
\Require \texttt{ncu\_rows}: list of \texttt{NcuRow} sorted by invocation index per kernel name
\Ensure  Each \texttt{KernelRow} assigned a \texttt{KernelMetrics} object
\State $\textit{counter} \gets \text{defaultdict}(\text{int})$
  \Comment{per kernel name}
\For{$k \in \texttt{nsys\_kernels}$}
  \State $i \gets \textit{counter}[k.\textit{name}]$
  \State $k.\textit{metrics} \gets \texttt{ncu\_rows}[k.\textit{name}][i]$
  \State $\textit{counter}[k.\textit{name}] \mathrel{+}= 1$
\EndFor
\end{algorithmic}
\end{algorithm}

This matching is correct when the workload is deterministic and input shapes are identical between
the \nsys capture and the \ncu replay.  Shape validation before the replay step enforces this
invariant (\secref{sec:edgecases}).  \figref{fig:iom} illustrates why timestamp joins fail and how
invocation-order matching avoids the problem.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\small,
  arr/.style={-{Stealth[length=4pt]}, thick},
  karr/.style={-{Stealth[length=4pt]}, thick, red!70!black},
  garr/.style={-{Stealth[length=4pt]}, thick, green!50!black},
  kern/.style={draw, rounded corners=2pt, fill=blue!10, draw=blue!60!black,
               minimum width=1.6cm, minimum height=1.4em, align=center, font=\footnotesize},
  kernncu/.style={draw, rounded corners=2pt, fill=purple!10, draw=purple!60!black,
               minimum width=1.6cm, minimum height=1.4em, align=center, font=\footnotesize},
  tl/.style={-{Stealth[length=5pt]}, thick, gray},
  lbl/.style={font=\scriptsize, text=gray!80!black},
  hl/.style={font=\footnotesize\bfseries},
  darr/.style={-{Stealth[length=4pt]}, thick, dashed, gray!50},
]

%% ---- (a) Naive timestamp join ----
\node[hl] at (4.5, 0.7) {(a) Naive timestamp join --- \textcolor{red!70!black}{incorrect}};

\draw[tl] (0.0, 0) -- (9.8, 0) node[right,lbl] {nsys (GPU clock)};
\node[lbl, left=2pt] at (0.0, 0) {\small nsys};

\node[kern] (k1n) at (1.1, 0)  {$K_1$\;\tiny t=10};
\node[kern] (k2n) at (4.0, 0)  {$K_2$\;\tiny t=45};
\node[kern] (k3n) at (7.1, 0)  {$K_3$\;\tiny t=80};

\draw[tl] (0.0,-1.25) -- (9.8,-1.25) node[right,lbl] {ncu (replay clock)};
\node[lbl, left=2pt] at (0.0,-1.25) {\small ncu};

\node[kernncu] (k1c) at (2.3,-1.25)  {$K_1$\;\tiny t=35};
\node[kernncu] (k2c) at (5.1,-1.25)  {$K_2$\;\tiny t=62};
\node[kernncu] (k3c) at (8.1,-1.25)  {$K_3$\;\tiny t=95};

\draw[darr] (k1c.north) -- (k1n.south);
\draw[karr] (k2c.north) -- (k1n.south)
  node[near start, left=3pt, lbl, text=red!70!black] {\textbf{\ding{55}} wrong match};
\draw[darr] (k3c.north) -- (k3n.south);

\draw[decorate,decoration={brace,amplitude=5pt},gray!60]
  (k1n.south east)++(0.1,0) -- (k1c.north east)++(0.1,0)
  node[midway, right=6pt, lbl, align=left] {clock\\drift};

%% ---- (b) IOM ----
\node[hl] at (4.5,-2.15) {(b) Invocation-order matching (IOM) --- \textcolor{green!50!black}{correct}};

\draw[tl] (0.0,-2.9) -- (9.8,-2.9) node[right,lbl] {nsys order};
\node[lbl, left=2pt] at (0.0,-2.9) {\small nsys};
\node[kern] (k1ni) at (1.4,-2.9) {$K_1$\;\tiny $i{=}0$};
\node[kern] (k2ni) at (4.0,-2.9) {$K_2$\;\tiny $i{=}1$};
\node[kern] (k3ni) at (6.7,-2.9) {$K_3$\;\tiny $i{=}2$};

\draw[tl] (0.0,-4.1) -- (9.8,-4.1) node[right,lbl] {ncu order};
\node[lbl, left=2pt] at (0.0,-4.1) {\small ncu};
\node[kernncu] (k1ci) at (1.4,-4.1) {$K_1$\;\tiny $i{=}0$};
\node[kernncu] (k2ci) at (4.0,-4.1) {$K_2$\;\tiny $i{=}1$};
\node[kernncu] (k3ci) at (6.7,-4.1) {$K_3$\;\tiny $i{=}2$};

\draw[garr] (k1ci.north) -- (k1ni.south) node[midway,right=1pt,lbl,text=green!50!black] {\textbf{\ding{51}}};
\draw[garr] (k2ci.north) -- (k2ni.south) node[midway,right=1pt,lbl,text=green!50!black] {\textbf{\ding{51}}};
\draw[garr] (k3ci.north) -- (k3ni.south) node[midway,right=1pt,lbl,text=green!50!black] {\textbf{\ding{51}}};

\node[lbl, align=center] at (4.9,-4.75)
  {Match key: \texttt{(kernel\_name, invocation\_index)} --- no timestamp comparison required};

\end{tikzpicture}
\caption{\textbf{Why timestamp joins fail and how IOM avoids them.}
  (a)~The nsys GPU-domain clock and the \ncu replay clock diverge by $>$10\,$\mu$s; a
  nearest-timestamp join misattributes $K_2$'s hardware counters to $K_1$.
  (b)~IOM matches kernels by ordinal position within each \texttt{(kernel\_name,
  invocation\_index)} sequence, making attribution clock-immune by construction.}
\label{fig:iom}
\end{figure}

\subsection{StreamIntervalTree Attribution}
\label{sec:intervaltree}

For each unique \texttt{(stream\_id, device\_id)} pair observed in the \nsys trace, \sys
constructs a \texttt{StreamIntervalTree}: a sorted list of NVTX ranges indexed by
\texttt{start\_ns}, implemented using Python's \texttt{sortedcontainers.SortedList} for
$O(\log n)$ insertion and $O(k \log n)$ enclosure queries ($k$ = number of enclosing ranges).

\begin{algorithm}[t]
\caption{NVTX Enclosure Attribution (MEDIUM Confidence)}
\label{alg:nvtx}
\begin{algorithmic}[1]
\Require Kernel row $k$ with \texttt{cpu\_launch\_start\_ns}, \texttt{stream\_id}
\Require \texttt{forest}: map from \texttt{(stream\_id, device\_id)} to \texttt{SortedList}
\Ensure Attribution at MEDIUM confidence, or fall through to heuristic
\State $T \gets k.\textit{cpu\_launch\_start\_ns}$
\State $\textit{ranges} \gets$ \texttt{forest}[$k.\textit{stream\_id}$].query\_enclosing($T$)
  \Comment{all NVTX ranges $[s,e]$ with $s \le T \le e$}
\If{$|\textit{ranges}| = 0$}
  \State \textbf{fall through} to Inductor fusion enrichment (\secref{sec:heuristic}) or UNATTRIBUTED
\ElsIf{$|\textit{ranges}| = 1$}
  \State $k.\textit{operator} \gets \textit{ranges}[0].\textit{text}$,
         \textit{confidence} $\gets$ MEDIUM
\Else
  \State $k.\textit{operator} \gets \arg\max_r r.\textit{nesting\_level}$
         \Comment{innermost (most specific) enclosing range}
  \State $k.\textit{is\_fused} \gets \top$;
         $k.\textit{all\_enclosing} \gets \textit{ranges}$
         \Comment{fused kernel spans multiple operators}
\EndIf
\end{algorithmic}
\end{algorithm}

The \texttt{cpu\_launch\_start\_ns} field is used as the query point rather than the GPU
\texttt{start\_ns}, because NVTX ranges are CPU-side events and the async gap between host launch
and GPU execution makes GPU timestamps unreliable for enclosure queries.

\subsection{Inductor Fusion Enrichment}
\label{sec:heuristic}

When NVTX enclosure fails (no enclosing range exists at the query point), \sys applies a
post-attribution enrichment pass that recovers ground-truth fused operator identity from Inductor
debug artifacts, rather than guessing from kernel names.

\textbf{Inductor debug extraction.}  When \texttt{torch.compile(backend="inductor")} runs with
debug mode enabled, Inductor writes \texttt{output\_code.py} artifacts containing the generated
Triton kernel source.  Each kernel call site is preceded by a structured comment:

\begin{lstlisting}[style=pystyle]
# Topologically Sorted Source Nodes: [...],
# Original ATen: [aten.relu, aten.addmm]
def triton_poi_fused_relu_addmm_0(in_ptr0, ...):
    ...
\end{lstlisting}

\sys's \texttt{InductorFusionExtractor} parses these comments via the regex
\texttt{\# .+Original~ATen:\textbackslash{}s*\textbackslash{}[([\textasciicircum{}\textbackslash{}]]+)\textbackslash{}]},
producing a mapping $\{$\textit{kernel\_short\_name} $\to$ $[$\texttt{aten::op}_1,
\texttt{aten::op}_2, $\ldots]$$\}$ that contains the exact set of aten operators fused into
each kernel.

\textbf{Enrichment as a post-attribution pass.}  This map is applied \emph{after} the NVTX
enclosure pass and never overrides a valid TORCH\_PROFILER or NVTX attribution:
\begin{enumerate}[leftmargin=1.5em]
  \item For kernels that remain UNATTRIBUTED after NVTX enclosure, a lookup in the fusion map
    upgrades them to \texttt{INDUCTOR\_FUSION} attribution at MEDIUM confidence, with
    \texttt{source\_operators} populated from the \texttt{Original~ATen} comment.
  \item For already-attributed kernels (TORCH\_PROFILER or NVTX), the enrichment augments the
    \texttt{fused\_ops} field and sets \texttt{is\_fused~=~True} if multiple operators are present.
\end{enumerate}

Kernels matching neither NVTX enclosure nor the Inductor fusion map remain in
\texttt{unattributed\_kernels[]} with \texttt{attribution\_method = "unattributed"}.  For eager-mode
workloads or cuDNN-backed operators (\eg \texttt{nn.LSTM} via \texttt{aten::\_cudnn\_rnn}), which
produce neither NVTX brackets nor Inductor debug output, unattributed rates of 20--85\% are expected
and are surfaced explicitly in the profile output.

\subsection{Confidence Tiers}
\label{sec:tiers}

\begin{table}[t]
\centering
\caption{Attribution methods in \sys.  Coverage percentages are measured over kernel
  \emph{runtime} (not kernel count); the UNATTRIBUTED rate varies widely with compile mode and
  operator backend.  Ranges shown are typical across PyTorch-native workloads; see
  \tabref{tab:coverage} for per-workload values.}
\label{tab:tiers}
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{p{2.0cm}p{2.6cm}p{4.8cm}p{1.6cm}}
\toprule
\textbf{Method} & \textbf{Evidence} & \textbf{Trigger Condition} & \textbf{Coverage} \\
\midrule
\rowcolor{highcol!15}
HIGH & torch.profiler IOM & Kineto trace available; kernel count agrees with dispatch count & 35--45\% \\
\rowcolor{medcol!15}
MEDIUM (NVTX) & NVTX enclosure (interval tree) & Kernel launch falls within an \texttt{aten::} NVTX range & 45--55\% \\
\rowcolor{medcol!10}
MEDIUM (Inductor) & Inductor \texttt{output\_code.py} comments & Kernel name in Inductor fusion map; NVTX enclosure absent & 5--15\% \\
\rowcolor{ucol!15}
UNATTRIBUTED & None & cuDNN-internal kernels (LSTM, cuDNN conv.); eager Triton without NVTX & 1--85\% \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Edge Cases}
\label{sec:edgecases}

\sys explicitly handles the following attribution edge cases:

\begin{enumerate}[leftmargin=1.5em]

\item \textbf{Clock domain mismatch.} Invocation-order matching (\algref{alg:iom}) makes
  attribution clock-immune by construction.

\item \textbf{CUDA Graph replay.} During graph capture, operators emit NVTX ranges and kernels are
  attributed normally.  During replay, new GPU timestamps with no NVTX correspondence are generated.
  \sys detects replay by the absence of NVTX events in the replay window and replays the attribution
  cache from the capture phase.

\item \textbf{Multi-stream execution.} Each \texttt{(stream\_id, device\_id)} pair has its own
  \texttt{StreamIntervalTree}, preventing cross-stream false matches.

\item \textbf{JIT warm-up kernels.} Kernels launched before the first NVTX range (JIT compilation
  artifacts) are flagged using a duration-outlier heuristic ($> 10\times$ median kernel duration)
  and excluded from operator mapping.

\item \textbf{Async kernel launch.} NVTX push/pop occurs on the host before the kernel executes on
  the GPU.  The RUNTIME table's \texttt{cpu\_launch\_start\_ns} field provides a host-side anchor
  that lies within the NVTX range, whereas the GPU \texttt{start\_ns} may not.

\item \textbf{Dynamic shapes.} \sys validates that input shapes at \ncu replay match those recorded
  during \nsys capture; a \texttt{ShapeMismatchError} is raised otherwise.

\item \textbf{Fused kernels.} When multiple NVTX ranges enclose a single kernel (Triton-fused
  cross-operator), \sys sets \texttt{is\_fused=True} and stores all enclosing ranges in
  \texttt{all\_enclosing\_ranges} for downstream analysis.

\item \textbf{ncu timing independence.} \ncu timestamps are internal to each replay run and are
  discarded entirely; only metric values are extracted from \ncu output.

\end{enumerate}

% -------------------------------------------------------
\section{Metric Set and Aggregation}
\label{sec:metrics}

\subsection{Counter Selection Rationale}

Collecting all 90+ \ncu counters is expensive (multiple replay passes) and produces profiles too
large for operator-level analysis.  We select 20 counters according to three criteria:
(1)~coverage of every bottleneck axis in the roofline taxonomy (\secref{sec:roofline});
(2)~availability of a valid counter name on Ampere, Hopper, \emph{and} Blackwell (some counters
were renamed in the Blackwell GB202 silicon); (3)~non-redundancy (no two selected counters
correlate perfectly).

\subsection{The 20-Counter Set}

\begin{table}[t]
\centering
\caption{The 20 hardware counters collected by \sys.  ``Agg.'' denotes the aggregation method
  applied when multiple kernels are attributed to one operator (see \secref{sec:aggregation}).}
\label{tab:counters}
\scriptsize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{clllc}
\toprule
\textbf{\#} & \textbf{Profile Field} & \textbf{Bottleneck Axis} & \textbf{Measures} & \textbf{Agg.}\\
\midrule
1  & \texttt{gpu\_time\_duration\_ns}      & Latency          & GPU wall time                & SUM \\
2  & \texttt{dram\_bytes\_read}            & Memory BW        & DRAM read bytes              & SUM \\
3  & \texttt{dram\_bytes\_written}         & Memory BW        & DRAM write bytes             & SUM \\
4  & \texttt{memory\_throughput\_pct}      & Memory BW        & Memory subsystem \% peak     & DW-mean \\
5  & \texttt{dram\_throughput\_pct}        & Memory BW        & DRAM \% of peak BW           & DW-mean \\
6  & \texttt{mem\_busy\_pct}              & L1 pipeline      & L1 pipeline \% peak active   & DW-mean \\
7  & \texttt{l1\_hit\_rate}               & Cache            & L1 cache hit rate (\%)       & DW-mean \\
8  & \texttt{l2\_hit\_rate}               & Cache            & L2 cache hit rate (\%)       & DW-mean \\
9  & \texttt{sm\_throughput\_pct}         & Compute          & SM \% of peak throughput     & DW-mean \\
10 & \texttt{tensor\_core\_active\_pct}   & Tensor Cores     & TC pipeline active (\%)      & DW-mean \\
11 & \texttt{sm\_active\_cycles}          & Compute          & Total SM active cycles       & SUM \\
12 & \texttt{achieved\_occupancy}         & Occupancy        & Warps active \% of peak      & DW-mean \\
13 & \texttt{warp\_cycles\_per\_instr}    & Latency          & Cycles per issued instruction & DW-mean \\
14 & \texttt{eligible\_cycles\_pct}       & Latency          & Eligible warp issue rate (\%) & DW-mean \\
15 & \texttt{executed\_instructions}      & Throughput       & Total executed instructions  & SUM \\
16 & \texttt{ipc\_active}                 & Throughput       & IPC when SM active           & DW-mean \\
17 & \texttt{avg\_threads\_per\_warp}     & Divergence       & Avg threads active per warp  & DW-mean \\
18 & \texttt{registers\_per\_thread}      & Reg.\ pressure   & Registers allocated / thread & MAX \\
19 & \texttt{local\_memory\_spills}       & Reg.\ spills     & Spill events to L1 local mem & SUM \\
20 & \texttt{dynamic\_smem\_per\_block}   & SMEM pressure    & Dynamic shared mem / block   & MAX \\
\bottomrule
\end{tabular}
\end{table}

\tabref{tab:counters} lists all 20 counters, their bottleneck axis, what they measure, and their
aggregation method.  Seven counters require architecture-specific name fallbacks between Ampere
(e.g.\ \texttt{dram\_\_bytes\_read.sum}), Hopper, and Blackwell
(e.g.\ \texttt{dram\_\_bytes\_op\_read.sum} or \texttt{fbpa\_\_dram\_read\_bytes.sum});
\sys queries all variants and uses the first that returns a value.

\subsection{Duration-Weighted Aggregation}
\label{sec:aggregation}

An operator may dispatch multiple kernels with heterogeneous durations.  For rate and utilization
metrics (rows marked DW-mean in \tabref{tab:counters}), a naive arithmetic mean would give a
short-lived auxiliary kernel the same weight as the long-running compute kernel.  \sys uses
duration-weighted aggregation instead:
\begin{equation}
  \overline{m}_\text{op} = \frac{\displaystyle\sum_{i} m_i \cdot d_i}{\displaystyle\sum_{i} d_i}
  \label{eq:dwmean}
\end{equation}
where $m_i$ and $d_i$ are the metric value and duration of the $i$-th kernel attributed to the
operator.  This ensures that the dominant kernel---the primary optimization target---contributes
proportionally to the operator-level summary.  For additive quantities (bytes, instruction counts,
spills), \sys uses plain summation.  For per-kernel constants (register count, shared memory),
\sys takes the maximum across all kernels, since the highest-pressure kernel limits occupancy.

% -------------------------------------------------------
\section{Profile-Guided Optimization Backend}
\label{sec:pgo}

\subsection{FX Pass Infrastructure}

\sys provides two composable components for graph-level optimization.
\textbf{UniqueSubgraphRegistry} partitions the FX graph by detected layer
structure, identifies structurally identical subgraphs via hash comparison,
and compiles a single representative per equivalence class.  Duplicate
partitions share the compiled callable, reducing compilation and ncu replay
time by $O(\text{num\_duplicate\_layers})$.

\textbf{FxPassRunner} applies user-defined pattern rewrites to unique
representatives only, then propagates the transformed compiled callables to
all structural duplicates:

\begin{lstlisting}[style=pystyle]
runner = FxPassRunner(registry)
for pattern_fn, replacement_fn in get_passes():
    runner.apply_pass(pattern_fn, replacement_fn)
\end{lstlisting}

Custom backends are registered via Torch Dynamo's backend hook and receive
the Aten IR \texttt{GraphModule} before Inductor lowering:

\begin{lstlisting}[style=pystyle]
@torch._dynamo.register_backend("my_workload_opt")
def my_backend(
    gm: torch.fx.GraphModule,
    example_inputs: list[torch.Tensor],
) -> Callable:
    # apply profile-guided modifications, then delegate to Inductor
    return compile_fx(gm, example_inputs,
                      config_patches={"max_autotune": True})
\end{lstlisting}

\subsection{Optimization Patterns}
\label{sec:passes}

Hardware-evidence-motivated optimization patterns are implemented per-workload
in each example's \texttt{*\_optimized.py}.  Patterns fall into two categories:
\emph{pre-compile model modifications} applied before \texttt{torch.compile} is
called, and \emph{FX graph pattern rewrites} applied inside a custom backend via
\texttt{torch.fx.subgraph\_rewriter.replace\_pattern}.  Each pattern fires only
when the profiler has diagnosed the specific counter symptom that motivates it.
\tabref{tab:passes} summarizes the patterns demonstrated across the evaluated
workloads.

\begin{table}[t]
\centering
\caption{Hardware-evidence-motivated optimization patterns demonstrated across
  the evaluated workloads.  ``Pre-compile'' patterns modify the \texttt{nn.Module}
  or tensor before \texttt{torch.compile}; ``FX rewrite'' patterns modify the
  Aten IR graph inside a custom backend.}
\label{tab:passes}
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{p{2.0cm}p{1.5cm}p{2.8cm}p{3.3cm}}
\toprule
\textbf{Pattern} & \textbf{Type} & \textbf{Counter Evidence} & \textbf{Transformation} \\
\midrule
BF16 promotion
  & Pre-compile
  & \texttt{tensor\_core\_active\_pct < 5\%}
  & \texttt{model.to(bfloat16)} before compile; routes GEMM to HMMA Tensor Core path \\[4pt]
\texttt{channels\_last} layout
  & Pre-compile
  & \texttt{convertTensor\_kernel} in \nsys trace
  & \texttt{model.to(channels\_last)}; eliminates cuDNN NCHW$\to$NHWC coercion \\[4pt]
BatchNorm folding
  & Pre-compile
  & \texttt{\_native\_batch\_norm\_legit\_no\_training} in profile
  & Fold BN running statistics into preceding Conv2d weights; replace module with \texttt{Identity} \\[4pt]
Weight pretranspose
  & FX rewrite
  & Low \texttt{sm\_throughput\_pct} on TN GEMM
  & Replace \texttt{aten.t(W)} + \texttt{aten.mm} with pre-computed \texttt{W$^\top$.contiguous()} buffer \\[4pt]
max-autotune
  & Compiler setting
  & Sub-optimal tile selection (SM throughput below peak)
  & \texttt{config\_patches=\{`max\_autotune': True\}} in \texttt{compile\_fx} \\
\bottomrule
\end{tabular}
\end{table}

\textbf{BF16 promotion} is the highest-leverage pattern.  On Ampere and newer
architectures, BF16 operands route all GEMM calls through Tensor Core HMMA tiles,
delivering up to $16\times$ raw FLOP throughput over FP32 scalar SIMT on A100
(312\,TFLOPS vs.\ 19.5\,TFLOPS).  It is applied only when the profiler diagnoses
\texttt{tensor\_core\_active\_pct} $\approx 0$, ruling out models that already use
mixed precision.

\textbf{Weight pre-transposition} uses \texttt{torch.fx.subgraph\_rewriter.replace\_pattern}
to detect the pattern \texttt{mm(x, aten.t(W))} where $W.\textit{shape}[0] \ge 512$.
It pre-computes $W^\top.\texttt{contiguous()}$ at graph compile time, registers it
as a model buffer, and replaces the runtime transpose node with a direct
\texttt{get\_attr}.  This switches cuBLAS from the \texttt{gemmSN\_TN} path
(sub-optimal L2 tiling for tall-$K$ TN matrices) to the \texttt{gemmSN\_NN} path.
In \texttt{gpt2\_optimized.py}, this pass ran but did not apply because the
\texttt{aten.t(get\_attr)} pattern is absent from the pre-Inductor graph; the pass
degrades gracefully with a diagnostic warning.

\subsection{Layer Deduplication}
\label{sec:dedup}

For models with repeated structure (\eg transformer decoder blocks), profiling and
compiling every layer independently costs $O(N)$ ncu replay time.
\texttt{UniqueSubgraphRegistry} detects structurally identical FX subgraphs by
computing a canonical hash of each partition's node types and connectivity.  Only
one representative per equivalence class is compiled and profiled; metrics and
compiled callables are propagated to all structural duplicates by positional index
within each partition group.

The deduplication is transparent to the outer profiling loop: each duplicate
partition's forward pass is wrapped with an NVTX range tagged
\texttt{layer::duplicate::<label>}, and the partition equivalence map (written to
\texttt{<prefix>.part.json} during the correlation pass) is passed to
\texttt{KernelProfileConfig} so the \ncu orchestrator skips duplicate-partition
kernels during replay and propagates metrics from the unique representative.

\textbf{Impact.}  For GPT-2 (12 identical transformer blocks),
deduplication reduces \ncu replay time from ${\sim}$30--45\,min to
${\sim}$3--5\,min---a 12$\times$ improvement.  Layer deduplication is activated
automatically by \texttt{run\_workload.py} when structurally repeated partitions
are detected.

\subsection{Closed-Loop Validation Protocol}
\label{sec:validation}

The full optimization cycle consists of four steps:
\begin{enumerate}
  \item \textbf{Profile}: capture baseline \texttt{profile.json} via \nsys + \ncu pipeline.
  \item \textbf{Diagnose}: \texttt{profile.json} is examined per-operator; the Claude Code
    plugin's \textbf{optimization-strategist} agent derives a time budget and edge case flags
    directly from \texttt{profile.json}, then reasons open-endedly from raw hardware counters
    to produce \texttt{optimizations.json} with ranked proposals and recommended patterns.
  \item \textbf{Transform}: implement a workload-specific optimized backend (or use the
    plugin's \textbf{backend-engineer} agent to generate one); run
    \texttt{torch.compile(model, backend="<workload>\_opt")} to apply the patterns and compile.
  \item \textbf{Validate}: capture optimized \texttt{profile\_optimized.json}; generate a delta
    report comparing per-operator counter values, flagging any regression.
\end{enumerate}

This protocol makes every optimization auditable: the delta report shows not just speedup, but
\emph{which counter improved}, providing evidence that the diagnosed bottleneck was actually
eliminated rather than merely shifted to a different operator.

\subsection{Automated Optimization via Claude Code Plugin}
\label{sec:plugin}

The closed-loop protocol above requires coordinating five distinct tools: \nsys, \ncu, the
attribution engine, the Torch Dynamo backend, and a second profiling pass.  \sys ships a
Claude Code plugin that automates all seven stages---baseline capture, FX
optimization proposals, backend code generation, 5-step validation, re-profiling,
cross-profile comparison, and human-readable reporting---in a single command:

\begin{lstlisting}[style=pystyle]
/optimize workload.py
\end{lstlisting}

The plugin orchestrates five specialized LLM agents.  \textbf{capture-agent} handles the \nsys
+ \ncu pipeline including executable auto-detection and \texttt{sudo} permission propagation;
\textbf{optimization-strategist} reads \texttt{profile.json} directly, derives a time budget
and edge case flags, and produces \texttt{optimizations.json}
with ranked FX proposals, exact metric-evidence citations, and dependency ordering;
\textbf{backend-engineer} generates \texttt{workload\_optimized.py} with workload-specific
optimization patterns (pre-compile modifications and/or FX graph rewrites) and a pytest suite; \textbf{validation-agent} runs the 5-step pass/fail validation sequence
and classifies which FX passes applied vs.\ fell through gracefully; and
\textbf{/report} normalizes for batch-size differences, attributes measured counter
deltas to specific transformations, and writes \texttt{report.md}.

Four Claude Code hooks maintain workflow continuity: write of \texttt{profile.json} prompts
\texttt{/propose}; write of \texttt{optimizations.json} prompts \texttt{/backend}; write of
\texttt{*\_optimized.py} prompts \texttt{/validate}; and a pre-Bash hook enforces
\texttt{--script-args} ordering in all \texttt{operator-profiler map} invocations.  A bundled
knowledge base provides GPU roofline reference data for 12 GPU architectures, canonical FX
pass implementations, and attribution edge-case guidance.  The plugin is installable without
cloning the repository:

\begin{lstlisting}[style=pystyle]
/plugin marketplace add yottalabsai/Profiler
/plugin install nvidia-profiler@profiler-plugins
\end{lstlisting}
