\section{Related Work}
\label{sec:related}

\paragraph{GPU profiling infrastructure.}
\texttt{torch.profiler} and its Kineto backend \cite{kineto2021} record operator-level wall-clock
durations and Chrome traces, but expose no hardware performance counters. NVIDIA Nsight
Systems~\cite{nsightsystems} and Nsight Compute~\cite{nsightcompute} together provide the data
that \sys consumes---timeline annotations and per-kernel hardware metrics---but provide no
automatic operator attribution: the bridge between the two tools has historically been manual.
NVIDIA DCGM and Dynolog \cite{dynolog2023} target cluster-level monitoring rather than
per-operator hardware attribution. \sys is the first system to \emph{automatically} attribute
\ncu hardware counters to \texttt{aten::} operators via NVTX enclosure and to close the
attribution-to-optimization loop within the Torch Dynamo compilation stack.

\paragraph{Deep learning compilers.}
TorchDynamo and TorchInductor~\cite{pytorch2024} compile PyTorch programs to Triton kernels using
heuristic fusion rules (pointwise chaining, matmul epilogue fusion), but do not incorporate
hardware counter feedback in pass selection.  \sys wraps Inductor as a downstream delegate and
inserts hardware-evidence-motivated FX passes \emph{before} Inductor's lowering step, creating a
feedback loop that Inductor alone lacks.  TVM~\cite{tvm2018} and Halide~\cite{halide2013} optimize
tensor programs via measurement-based auto-tuning and schedule/algorithm separation respectively;
neither uses per-kernel hardware counters to decide which optimization class to apply.
ONNX Runtime and TensorRT perform layer fusion and quantization but present black-box interfaces
that cannot be extended for custom architectures or augmented with profiling evidence.

\paragraph{Attention and fused kernel libraries.}
FlashAttention~\cite{flashattention2022} and FlashAttention-2~\cite{flashattention2023} tile
attention computation to keep the $[B, H, S, S]$ attention matrix in on-chip SRAM, eliminating
DRAM round-trips and the register-spill pathology diagnosed by \sys in the SDPA workload.
\sys's FlashAttention substitution pass identifies when profiling evidence (\texttt{l2\_hit\_rate
$< 20\%$}, local memory spills $> 0$) warrants the substitution and applies it automatically via
an FX graph pattern replacement. xFormers~\cite{xformers2022} provides a static library of fused
attention variants; \sys is complementary: it determines \emph{which} variant is warranted for a
specific operator's hardware counter profile. Triton~\cite{triton2019} provides the JIT compilation
framework underlying Inductor kernels; \sys additionally \emph{parses} Triton kernel names to
recover fused operator identity when NVTX annotation is unavailable.

\paragraph{Profile-guided optimization.}
Classic PGO in C/C++ compilers (GCC, LLVM) collects branch frequency profiles and uses them to
guide inlining, unrolling, and layout decisions \cite{autofdo2016}.  Our system is the neural
network analog: instead of branch counts, we collect hardware performance counter profiles and use
them to select FX graph transformations.  The roofline model~\cite{williams2009roofline}---the
theoretical framework underlying our bottleneck taxonomy---has been applied to HPC kernels but not
previously operationalized for automated operator-level pass selection in a deep learning compiler.

\paragraph{Inference optimization frameworks.}
vLLM~\cite{vllm2023} and SGLang~\cite{sglang2024} achieve high LLM serving throughput via
PagedAttention, continuous batching, and memory pooling.  These frameworks operate at the
scheduling level; their hand-coded kernels (BF16 GEMMs, FA2) represent the operator-level
optimizations that \sys \emph{discovers automatically} from hardware profiling for arbitrary
architectures.  The two systems are complementary: \sys can profile a workload running inside
vLLM to identify residual kernel-level inefficiencies not covered by serving-layer scheduling.
